Generating Verified LLVM from Isabelle/HOL

### 2019-07-02 paper/main.tex
bounds checker

\author{Peter Lammich}{The University of Manchester}{peter.lammich@manchester.ac.uk}{}{}
\authorrunning {P. Lammich}

@@

@@

This paper presents a next step towards efficient verified algorithms: A refinement framework to generate verified code in LLVM intermediate representation~\cite{LLVM-manual} 
with total correctness guarantees. LLVM is an imperative intermediate language with a powerful and well-tested optimizing compiler.
We first formalize the semantics of Isabelle-LLVM, a simple imperative language shallowly embedded into Isabelle/HOL, 
and designed to be easily translated to actual LLVM text (\S\ref{sec:semantics}). 
On top of Isabelle-LLVM, we build a separation logic and a verification condition generator, which allows convenient reasoning about Isabelle-LLVM programs (\S\ref{sec:vcg}).
Finally, we modify the Sepref tool to target Isabelle-LLVM instead of Imperative/HOL (\S\ref{sec:auto_ref}), connecting 
the Refinement Framework to our LLVM code generator. This only affects the last refinement step, 
such that most parts of existing verifications can be reused. 
As case studies (\S\ref{sec:casestudies}), we verify a binary search algorithm and adopt an existing formalization~\cite{HeLa17} of
the Knuth-Morris-Pratt string search algorithm~\cite{KMP77}. The resulting LLVM code is significantly
faster than the corresponding Standard-ML code and on par with unverified C implementations.
The paper ends with the discussion of future work (\S\ref{sec:future_work}) and related work (\S\ref{sec:related_work}).
%
The Isabelle theories described in this paper are available at \url{http://www21.in.tum.de/~lammich/isabelle_llvm}.

@@

We use do-notation, i.e.\ \q{\is|do { x<-m; f x }|} is short for \q{\is{bind m (\<lambda>x. f x)}}.
Moreover, we define a flat chain complete partial order~\cite{Mark76} on \q{\is{mres}}, with \q{\is{\<bot> := NTERM}}.
For a monotonic function \q{\is{F :: ('a => ('b,'s) M) => 'a => ('b,'s) M}}, \q{\is{REC F}} is the least fixed point.
As functions defined using the monad combinators are monotonic by construction~\cite{Kr10}, 
we can define arbitrary recursive computations. The partial function package~\cite{Kr10} supports automation for monotonicity 
proofs and for defining simple recursive functions. Mutual recursion still requires some manual effort, though it could be automated, too.

@@

\subsection{Towards a Shallow Embedding}
While we explicitly model values in memory by the type \q{\is{val}}, we model values in registers in a more shallow fashion:
We identify LLVM registers with Isabelle variables that have a type of the shape \q{\is{T = T \x T | n word | T ptr}}. 
Here, \q{\is{\x}} is Isabelle's product type, \q{\is{n word}} is the $n$ bit word type from Isabelle's word 
library\cprotect\footnote{For convenient notation, we use the type \q{\is{n word}} as if it were a type depending on a variable $n$. 
Isabelle/HOL is not dependently typed. Instead, $n$ is actually a type variable with type-class \q{\is{len}}, which provides a 
function \q{\is{len_of :: 'a::len itself => nat}} to extract the length as a term.}, 
and \q{\is{'a ptr}} is a pointer with an attached phantom type for the value pointed to (\q{\is{'a ptr = PTR rptr}}).
%
For each type \q{\is{'a}} of shape \q{\is{T}}, we define the functions:
\begin{lstlisting}
to_val     :: "'a => val"            struct_of    :: "'a itself => vstruct"
from_val :: "val => 'a"             init          :: 'a
\end{lstlisting}
such that
\begin{lstlisting}
"from_val o to_val = id"                "vstruct (to_val x) = (struct_of TYPE('a))"
"to_val init = zero_initializer (struct_of TYPE('a))"
\end{lstlisting}
Here, \q{\is{TYPE('a) :: 'a itself}} reflects type \q{\is{'a}} into a term.
The functions \q{\is{to_val}} and \q{\is{from_val}} inject a T-shaped type \q{\is{'a}} into a value with structure \q{\is{struct_of TYPE('a)}}.
Moreover, \q{\is{init::'a}} corresponds to the all-zeroes value, i.e., the value where all pointers are null pointers, and all integers are $0$.

@@

For example, the \q{\is{llc_while}} combinator is defined as a fixed point (cf.~\S\ref{sec:modeling_ctrl_flow}). 
Using its definition as an unfold rule, the preprocessor will automatically convert while loops into tail calls.
This allows for using while-loops without trusting their translation in the code generator. 
% However, to be efficient, we have to enable LLVM's tail call optimization, which will translate the function back into a loop.
A configuration option in our tool lets the user choose between direct while-loop translation or unfolding to a tail call.

@@

\subsection{Basic Rules}
Once we have set up the separation algebra and the abstraction function, we can prove Hoare triples for the basic operations of our memory model.
For example, we prove the following rules for \q{\is{allocn}} and \q{\is{free}}:
\begin{lstlisting}
|= {\<box>} allocn v n {%p. malloc_tag n p * range {0..<n} (%_. v) p}
|= {malloc_tag n p * \<exists>blk. range {0..<n} blk p} free p {%_. \<box>}
\end{lstlisting}
where \q{\is{\<box> = \<lambda>s. s=0}} describes the empty memory, 
\q{\is{malloc_tag n p}} asserts that \q{\is{p}} points to the beginning of a block, and the size field of this block's abstraction is \q{\is{n}}, 
and \q{\is{range I f p}} describes that for all \q{$i\in I$}, \q{$p+i$} points to value \q{$f~i$}.
Intuitively, \q{\is{allocn}} creates a block of size \q{$n$}, initialized with values \q{$v$}, and a tag. 
If one possesses both, the whole block and the tag, it can be deallocated by free. 
% Note that we use the size encoded in the tag to ensure that we possess the whole block. 
For the Isabelle-LLVM memory instructions, we obtain the following rules:
\begin{lstlisting}
|= {n~=0} ll_malloc TYPE('a) n {%p. range {0..<n} (%_. init) p * malloc_tag n p}
|= {range {0..<n} blk p * malloc_tag n p} ll_free p {%_. \<box>}
|= {pto x p} ll_load p {%r. r=x * pto x p}
|= {pto xx p} ll_store x p {%_. pto x p}
\end{lstlisting}
Here, \q{\is{pto x p}} describes that $p$ points to value $x$, and we write predicates as if they were assertions 
on the empty memory, e.g., \q{\is{r=x}} instead of \q{\is{\<lambda>s. s=0 \<and> r=x}}.
We prove similar rules for the other instructions.

@@

In separation logic based reasoning, a data structure provides a \emph{refinement assertion} \q{\is{A x x\impl :: assn}}, which describes 
that the abstract value \q{\is{x}} is implemented by the concrete value \q{\is{x\impl}}. 
We define refinement assertions to implement integers and natural numbers by $n$ bit words, and to implement lists by blocks of memory. 
Note that new data structures and refinement assertions can easily be added. In general, an implementation does not completely implement an abstract mathematical concept.
For example, $n$ bit words can only represent the integers \q{\is|sints n = $\{-2^{\text{n}-1}..<2^{\text{n}-1}\}$|}, and hash-tables can only represent finite sets. Thus, the rules for the operations generally come with additional preconditions. 
For example, the rule to implement subtraction on integers by subtraction on $n$ bit words is the following:
\begin{lstlisting}
|= {sint$_n$ a a\impl * sint$_n$ b b\impl * a-b \<in> sints n} ll_sub a\impl b\impl {%r\impl. sint$_n$ (a-b) r\impl}
  for a\impl b\impl :: n word and a b :: int
\end{lstlisting}
Here, \q{\is{sint$_n$}} implements mathematical integers by $n$-bit words.
Note that the postcondition does not mention the operands \q{\is{a,b}} again, though they are still valid after the operation. 
As \q{\is{sint$_n$}} is \emph{pure}, i.e., does not use the memory, our VCG will automatically add the corresponding assertions to the postcondition.

@@

We start with a high-level specification: For a list \q{\is{xs}} and a value \q{\is{x}}, find the index of the first element greater or equal to \q{\is{x}}.
We define the following constant:
\begin{lstlisting}
  fi_spec xs x = spec i. find_index (%y. x<=y) xs
\end{lstlisting}
where \q{\is{find_index P xs}} is a standard list function that returns the index of the 
first element in \q{\is{xs}} that satisfies \q{\is{P}}, or \q{\is{length xs}} if there is no such element.

@@

The Sepref tool implements mathematical integers by 64-bit words, proving absence of overflows.
This is only possible because the assertion in \q{\is{bin_search}} explicitly states that the indexes are in bounds. 
Moreover, note the expression \q{\is{l + (h-l) div 2}} that we used to compute the midpoint index. 
On mathematical integers, it is equal to \q{\is{(l+h) div 2}}. However, on fixed-width words, the latter may overflow, while the former does not\footnote{Exactly this 
overflowing midpoint computation caused the infamous bug in the Java Standard Library~\cite{bs_flaw_blogpost}.}.

@@

Our code generator, as well as most standard code generators in theorem provers, 
translate from logic to target language code, implicitly identifying logical 
concepts with programming language concepts. This approach is simple, however, 
the translation algorithm and its implementation become part of the trusted code base.
More recently, code generators that translate into a deeply embedded semantics of the 
target language have been developed~\cite{MyOw14,HuNi18}. 
We leave a translation to a deep embedding of LLVM to future work, and note that a 
deep embedding will also enable more advanced control flow constructs like exceptions and breaking from loops, 
without significantly increasing the trusted code base.

Compared to actual LLVM, Isabelle-LLVM makes a few simplifying assumptions: It assumes an infinite supply of memory, and thus cannot assign a bit-size to pointers. 
This assumption helps us to retain a deterministic semantics, which is executable inside the theorem prover (cf.~Example~\ref{ex:fib_isabelle}). 
We plan to use this feature for systematic testing of our code generator against the actual LLVM compiler. 
A similar assumption is implicitly made for the stack, as our semantics permits arbitrarily deep recursive procedure calls.
We remedy this mismatch between semantics and reality by terminating the program in a defined way if it runs out of heap.
To protect against stack overflows, LLVM provides mechanisms like stack probing or split stack, which, however, require some effort to enable. 
We leave that to future work, and note that our generated code allocates no large blocks of memory on the stack. 
Thus, stack overflows are likely to hit the guard pages inserted by most operating systems, which will cause termination of the process.

@@

The Vellvm project~\cite{ZNMZ12,ZNMZ13} verifies LLVM program transformations in Coq.
To be useful, e.g. as backend for clang, they have to formalize a substantial fragment of LLVM. 
On the other hand, we can afford to formalize a simplified and abstract semantics that 
is just powerful enough to cover what Sepref generates.

>>>>

\author{Peter Lammich}{The University of Manchester}{peter.lammich@manchester.ac.uk}{}{}
\authorrunning {P. Lammich}
\Copyright{Peter Lammich}

@@

\supplement{\url{http://www21.in.tum.de/~lammich/isabelle_llvm}}

@@

This paper presents a next step towards efficient verified algorithms: A refinement framework to generate verified code in LLVM intermediate representation~\cite{LLVM-manual} 
with total correctness guarantees. LLVM is an imperative intermediate language with a powerful and well-tested optimizing compiler.
We first formalize the semantics of Isabelle-LLVM, a simple imperative language shallowly embedded into Isabelle/HOL, 
and designed to be easily translated to actual LLVM text (\S\ref{sec:semantics}). 
On top of Isabelle-LLVM, we build a separation logic and a verification condition generator, which allows convenient reasoning about Isabelle-LLVM programs (\S\ref{sec:vcg}).
Finally, we modify the Sepref tool to target Isabelle-LLVM instead of Imperative/HOL (\S\ref{sec:auto_ref}), connecting 
the Refinement Framework to our LLVM code generator. This only affects the last refinement step, 
such that most parts of existing verifications can be reused. 
As case studies (\S\ref{sec:casestudies}), we verify a binary search algorithm and adopt an existing formalization~\cite{HeLa17} of
the Knuth-Morris-Pratt string search algorithm~\cite{KMP77}. The resulting LLVM code is significantly
faster than the corresponding Standard-ML code and on par with unverified C implementations.
The paper ends with the discussion of future work (\S\ref{sec:future_work}) and related work (\S\ref{sec:related_work}).
%
The Isabelle theories described in this paper are available as supplement material (URL displayed in paper header).

@@

We use do-notation, i.e.\ \q{\is|do { x<-m; f x }|} is short for \q{\is{bind m (\<lambda>x. f x)}}.
Moreover, we define a flat chain complete partial order~\cite{Mark76} on \q{\is{mres}}, with \q{\is{\<bot> := NTERM}}.
For a monotonic function \q{\is{F :: ('a => ('b,'s) M) => 'a => ('b,'s) M}}, \q{\is{REC F}} is the least fixed point.
As functions defined using the monad combinators are monotonic by construction~\cite{Kr10}, 
we can define arbitrary recursive computations. The partial function package~\cite{Kr10} provides automation for monotonicity 
proofs and for defining simple recursive functions. Mutual recursion still requires some manual effort, though it could be automated, too.

@@

\subsection{Towards a Shallow Embedding}
While we explicitly model values in memory by the type \q{\is{val}}, we model values in registers in a more shallow fashion:
We identify LLVM registers with Isabelle variables that have a type of shape \q{\is{T = T \x T | n word | T ptr}}. 
Here, \q{\is{\x}} is Isabelle's product type, \q{\is{n word}} is the $n$ bit word type from Isabelle's word 
library\cprotect\footnote{For convenient notation, we use the type \q{\is{n word}} as if it were a type depending on a variable $n$. 
Isabelle/HOL is not dependently typed. Instead, $n$ is actually a type variable with type-class \q{\is{len}}, which provides a 
function \q{\is{len_of :: 'a::len itself => nat}} to extract the length as a term.}, 
and \q{\is{'a ptr}} is a pointer with an attached phantom type for the value pointed to (\q{\is{'a ptr = PTR rptr}}).
%
For each type \q{\is{'a}} of shape \q{\is{T}}, we define the functions:
\begin{lstlisting}
to_val     :: "'a => val"            struct_of    :: "'a itself => vstruct"
from_val :: "val => 'a"             init          :: 'a
\end{lstlisting}
such that
\begin{lstlisting}
"from_val o to_val = id"                "vstruct (to_val x) = (struct_of TYPE('a))"
"to_val init = zero_initializer (struct_of TYPE('a))"
\end{lstlisting}
Here, \q{\is{TYPE('a) :: 'a itself}} reflects type \q{\is{'a}} into a term.
The functions \q{\is{to_val}} and \q{\is{from_val}} inject a T-shaped type \q{\is{'a}} into a value with structure \q{\is{struct_of TYPE('a)}}.
Moreover, \q{\is{init::'a}} corresponds to the all-zeroes value, i.e., the value where all pointers are null pointers, and all integers are $0$.

@@

For example, the \q{\is{llc_while}} combinator is defined as a fixed point (cf.~\S\ref{sec:modeling_ctrl_flow}). 
Using its definition as an unfold rule, the preprocessor will automatically convert while loops into tail calls.
This allows for using while-loops without trusting their translation in the code generator. 
% However, to be efficient, we have to enable LLVM's tail call optimization, which will translate the function back into a loop.
A configuration option in our tool lets the user choose between direct while-loop translation or unfolding into a tail call.

@@

\subsection{Basic Rules}
Once we have set up the separation algebra and the abstraction function, we can prove Hoare triples for the basic operations of our memory model.
For example, we prove the following rules for \q{\is{allocn}} and \q{\is{free}}:
\begin{lstlisting}
|= {\<box>} allocn v n {%p. malloc_tag n p * range {0..<n} (%_. v) p}
|= {malloc_tag n p * \<exists>blk. range {0..<n} blk p} free p {%_. \<box>}
\end{lstlisting}
where \q{\is{\<box> = \<lambda>s. s=0}} describes the empty memory, 
\q{\is{malloc_tag n p}} asserts that \q{\is{p}} points to the beginning of a block, and the size field of this block's abstraction is \q{\is{n}}, 
and \q{\is{range I f p}} describes that for all \q{$i\in I$}, \q{$p+i$} points to value \q{$f~i$}.
Intuitively, \q{\is{allocn}} creates a block of size \q{$n$}, initialized with values \q{$v$}, and a tag. 
If one possesses both, the whole block and the tag, it can be deallocated by free. 
% Note that we use the size encoded in the tag to ensure that we possess the whole block. 
For the Isabelle-LLVM memory instructions, we obtain the following rules:
\begin{lstlisting}
|= {n~=0} ll_malloc TYPE('a) n {%p. range {0..<n} (%_. init) p * malloc_tag n p}
|= {range {0..<n} blk p * malloc_tag n p} ll_free p {%_. \<box>}
|= {pto x p} ll_load p {%r. r=x * pto x p}
|= {pto xx p} ll_store x p {%_. pto x p}
\end{lstlisting}
Here, \q{\is{pto x p}} describes that $p$ points to value $x$, and we write predicates as if they were assertions 
on the empty memory, e.g., \q{\is{n~=0}} instead of \q{\is{\<lambda>s. s=0 \<and> n~=0}}.
We prove similar rules for the other instructions.

@@

In separation logic based reasoning, a data structure provides a \emph{refinement assertion} \q{\is{A x x\impl :: assn}}, which describes 
that the abstract value \q{\is{x}} is implemented by the concrete value \q{\is{x\impl}}. 
We define refinement assertions to implement integers and natural numbers by $n$ bit words, and to implement lists by blocks of memory.
On top of that, we define more complex data structures like dynamic arrays.
Note that new data structures can easily be added. In general, an implementation does not completely implement an abstract mathematical concept.
For example, $n$ bit words can only represent the integers \q{\is|sints n = $\{-2^{\text{n}-1}..<2^{\text{n}-1}\}$|}, and hash-tables can only represent finite sets. Thus, the rules for the operations generally come with additional preconditions. 
For example, the rule to implement subtraction on integers by subtraction on $n$ bit words is the following:
\begin{lstlisting}
|= {sint$_n$ a a\impl * sint$_n$ b b\impl * a-b \<in> sints n} ll_sub a\impl b\impl {%r\impl. sint$_n$ (a-b) r\impl}
  for a\impl b\impl :: n word and a b :: int
\end{lstlisting}
Here, \q{\is{sint$_n$}} implements mathematical integers by $n$-bit words.
Note that the postcondition does not mention the operands \q{\is{a,b}} again, though they are still valid after the operation. 
As \q{\is{sint$_n$}} is \emph{pure}, i.e., does not use the memory, our VCG will automatically add the corresponding assertions to the postcondition.

@@

We start with a high-level specification: For a list \q{\is{xs}} and a value \q{\is{x}}, find the index of the first element greater or equal to \q{\is{x}}.
We define the following constant:
\begin{lstlisting}
  fi_spec xs x = spec i. i = find_index (%y. x<=y) xs
\end{lstlisting}
where \q{\is{find_index P xs}} is a standard list function that returns the index of the 
first element in \q{\is{xs}} that satisfies \q{\is{P}}, or \q{\is{length xs}} if there is no such element.

@@

The Sepref tool implements mathematical integers by 64-bit words, proving absence of overflows.
This is only possible because the assertion in \q{\is{bin_search}} explicitly states that the indexes are in bounds. 
Moreover, note the expression \q{\is{l + (h-l) div 2}} that we used to compute the midpoint index. 
On mathematical integers, it is equal to \q{\is{(l+h) div 2}}. However, on fixed-width words, the latter may overflow, while the former does not\footnote{Exactly this 
overflow caused the infamous bug in the Java Standard Library~\cite{bs_flaw_blogpost}.}.

@@

Our code generator, as well as most standard code generators in theorem provers, 
translates from logic to target language code, implicitly identifying logical 
concepts with programming language concepts. This approach is simple, however, 
the translation algorithm and its implementation become part of the trusted code base.
More recently, code generators that translate into a deeply embedded semantics of the 
target language have been developed~\cite{MyOw14,HuNi18}. 
We leave a translation to a deep embedding of LLVM to future work, and note that a 
deep embedding will also enable more advanced control flow constructs like exceptions and breaking from loops, 
without significantly increasing the trusted code base.

Compared to actual LLVM, Isabelle-LLVM makes a few simplifying assumptions: 
We do not support floating point arithmetic, though this could be added, e.g.~based on Lei Yu's floating point formalization \cite{Yu13}.
Moreover, we only support two-element structures (pairs). This nicely fits Isabelle HOL's product datatype, and the nested structures 
resulting from longer tuples should not be a problem for LLVM's optimizer.
Also, we do not support concepts that are handy for program optimization, but not required for code generation, like poison values.
Isabelle-LLVM assumes an infinite supply of memory, and thus cannot assign a bit-size to pointers. 
This assumption helps us to retain a deterministic semantics, which is executable inside the theorem prover (cf.~Example~\ref{ex:fib_isabelle}). 
We plan to use this feature for systematic testing of our code generator against the actual LLVM compiler. 
A similar assumption is implicitly made for the stack, as our semantics permits arbitrarily deep recursive procedure calls.
We remedy this mismatch between semantics and reality by terminating the program in a defined way if it runs out of heap.
To protect against stack overflows, LLVM provides mechanisms like stack probing or split stack, which, however, require some effort to enable. 
We leave that to future work, and note that our generated code allocates no large blocks of memory on the stack. 
Thus, stack overflows are likely to hit the guard pages inserted by most operating systems, which will cause defined termination of the process.

@@

The Vellvm project~\cite{ZNMZ12,ZNMZ13} verifies LLVM program transformations in Coq.
To be useful, e.g.~as backend for clang, they have to formalize a substantial fragment of LLVM. 
On the other hand, we can afford to formalize a simplified and abstract semantics that 
is just powerful enough to cover what Sepref generates.
